The Concurrent Calculi Formalisation Benchmark

### 2024-02-02 short-paper/short-paper.tex
minor edits and rewrites

\subsection{Goals and how to contribute}
Similarly to the authors of the POPLMark challenges, we seek to
answer several questions:
\alb{We oscillate between concurrency in general and process calculi, here and before}
\begin{enumerate}[label=\textbf{(Q\arabic*)},leftmargin=10mm]
\item\label{item:rq1} What is the current state of the art in mechanising process calculi?
\item\label{item:rq2} Which techniques and best practices can be recommended when starting formalisation projects involving process calculi?
\item\label{item:rq3} Which improvements are needed to make mechanisation tools more user-friendly with regards to the issues faced when mechanising process calculi?
\end{enumerate}

@@

\subsubsection{General design remarks.}
Like the POPLMark and POPLMark Reloaded challenges, our challenge is not meant to be comprehensive:
applications such as multiparty session types~\cite{10.1145/2827695,10.1145/3290343}, choreographies~\cite{DBLP:journals/jar/CruzFilipeMP23}, conversation types~\cite{DBLP:journals/tcs/CairesV10}, psi-calculi~\cite{lmcs:696}, or encodings between different calculi~\cite{DBLP:journals/iandc/Gorla10,DBLP:conf/forte/CairesP16,DBLP:journals/iandc/DardhaGS17,DBLP:conf/ecoop/ScalasDHY17,DBLP:journals/iandc/KouzapasPY19,10.1145/3479394.3479407} are not directly covered.\alb{Do we need all those citations?}
Still, these (and other) applications need the basic techniques that our challenge problems emphasize: as per design goal \ref{item:goal-reusability}, our problems are drawn from the basic meta-theory of process calculi (without requiring the development of new theory), and focus on well-known theorems and results that involve interesting proof techniques that may be reused in further work.
To achieve design goal \ref{item:goal-comperison-accessibility},
we have formulated our challenge problems to explore the three aspects (linearity, scope extrusion, coinduction) independently, so that they may be solved individually and in any order;
each problem should be reasonably small and easily understood with basic knowledge of textbook concurrency theory, process calculi, and type theory.  The process calculus used in each challenge
focuses on the features that we want to emphasize, and omits all constructs
(such as choices) that would complicate the mechanisation without bringing additional tangible
benefits and insights. 
The minimality and uniformity of the calculi also allows us to target design
goal \ref{item:goal-tutorials}.

\subsubsection{Linearity.}
For our challenge on linear reasoning we have chosen a type preservation theorem
for a system using session types, whose use are common in concurrency theory. This allows us to introduce linear
reasoning with few definitions. Accordingly, we omit channel delegation and use a reduction instead of a
transition system semantics.
Inspired by Vasconcelos~\cite{Vasconcelos2012}, we define a syntax where
restriction binds two names, which highlights their duality within the type system.
We have chosen a simple notion of well-formedness,
and do not, for instance, consider deadlocks. While more
sophisticated notions of well-formedness are interesting, the
proofs of these properties would distract from the core linear aspects of the challenge.

@@

\subsubsection{Coinduction.}
Our coinduction challenge concerns strong barbed bisimulation and congruence.\alb{we motivate why strong vs weak, but not why barbed}
Though weak barbed congruence is a more common behavioural equivalence for \(\pi\)-terms,\alb{? why terms?} here we prefer strong equivalences to avoid the need to abstract over the number of internal transitions, thus simplifying the theory.
We exclude delegation from the coinduction challenge since it is orthogonal to our primary aim of exploring coinductive proof techniques.

@@

\paragraph{Extending and combining the current challenges.}
\todo{This paragraph is very long for how relatively little it says. Maybe remove/cut down?}
While we have tried to minimise the overlap of the techniques and aspects our challenges explore (linearity, scope extrusion, coinduction), most research on concurrency theory combines some or all of these.
A second round of formalisations might entail combining the features of the challenges.
We consider the following three extensions to be obvious places to start:
\emph{(1)} the addition of choice to the type level and the process level,
\emph{(2)} the addition of recursion and (for typed systems) recursive types,
and \emph{(3)} the addition of channel delegation to those calculi that do not already support it.%

\paragraph{Addressing new calculi and new features.}
\todo{Maybe cite some references in this section?}
Extending in this direction means proposing new challenges covering different features of process calculi, or different proof mechanisation features.
Some interesting aspects of message-passing calculi to be explored in further challenges could be multiparty session types and choreographies, as their meta-theory includes aspects -- \eg, well-formedness conditions on global types, partiality of end-point projection function, etc. -- that we do not address here.
Also, one could easily design a challenge that goes beyond our coinduction challenge, by exploring different notions of bisimilarity (barbed, weak, \etc) and trace equivalence.
%
\luca{Coinduction may also play a relevant role in supporting recursive
processes and session types. Indeed, it is often the case that session type
systems (and logics) with recursive session types (and propositions) are
naturally expressed in \emph{infinitary form} by interpreting their typing (and
proof) rules coinductively. I CAN ADD ONE OR TWO REFERENCES IF YOU FEEL IT'S APPROPRIATE}

>>>>

\subsection{Goals and how to contribute}
Similarly to the authors of the POPLMark challenges, we seek to
answer several questions:
%\alb{We oscillate between concurrency in general and process calculi, here and before}
\begin{enumerate}[label=\textbf{(Q\arabic*)},leftmargin=10mm]
\item\label{item:rq1} What is the current state of the art in mechanising process calculi?
\item\label{item:rq2} Which techniques and best practices can be recommended when starting formalisation projects involving process calculi?
\item\label{item:rq3} Which improvements are needed to make mechanisation tools more user-friendly with regards to the issues faced when mechanising process calculi?
\end{enumerate}

@@

\subsubsection{General design remarks.}
Like the POPLMark and POPLMark Reloaded challenges, our challenge is not meant to be comprehensive:
applications such as multiparty session types~\cite{10.1145/2827695,10.1145/3290343}, choreographies~\cite{DBLP:journals/jar/CruzFilipeMP23}, conversation types~\cite{DBLP:journals/tcs/CairesV10}, psi-calculi~\cite{lmcs:696}, or encodings between different calculi~\cite{DBLP:journals/iandc/Gorla10,DBLP:conf/forte/CairesP16,DBLP:journals/iandc/DardhaGS17,DBLP:conf/ecoop/ScalasDHY17,DBLP:journals/iandc/KouzapasPY19,10.1145/3479394.3479407} are not directly covered.\alb{Do we need all those citations about encodings?}
Still, these (and other) applications need the basic techniques that our challenge problems emphasize: as per design goal \ref{item:goal-reusability}, our problems are drawn from the basic meta-theory of process calculi (without requiring the development of new theory), and focus on well-known theorems and results that involve interesting proof techniques that may be reused in further work.
To achieve design goal \ref{item:goal-comperison-accessibility},
we have formulated our challenge problems to explore the three aspects (linearity, scope extrusion, coinduction) independently, so that they may be solved individually and in any order;
each problem should be reasonably small and easily understood with basic knowledge of textbook concurrency theory, process calculi, and type theory.  The process calculus used in each challenge
focuses on the features that we want to emphasize, and omits all constructs
(such as choices) that would complicate the mechanisation without bringing additional tangible
benefits and insights. 
The minimality and uniformity of the calculi also allows us to target design
goal \ref{item:goal-tutorials}.

\subsubsection{Linearity.}
For our challenge on linear reasoning we have chosen a type preservation theorem
for a system using session types.
% , whose use are common in concurrency theory.
This allows us to introduce linear reasoning with few definitions.
With our objective focused on linearity, we forego channel delegation
and opt for a reduction approach rather than employing a transition
system semantics.  Inspired by Vasconcelos~\cite{Vasconcelos2012}, we
define a syntax where restriction binds two names,n highlighting
their duality within the type system.  We have chosen a simple notion
of well-formedness, and do not  consider
deadlocks. While more sophisticated notions of well-formedness are
interesting, the proofs of these properties would distract from the
core linear aspects of the challenge.

@@

\subsubsection{Coinduction.}
Our coinduction challenge concerns strong barbed bisimulation and
congruence.\alb{we motivate why strong vs weak, but not why barbed}
Though weak barbed congruence is a more common behavioural equivalence
w.r.t.\ the \(\pi\)-calculus, here we prefer strong equivalences to avoid the
need to abstract over the number of internal transitions, thus
simplifying the theory.  We exclude delegation from the coinduction
challenge since it is orthogonal to our primary aim of exploring
coinductive proof techniques.

@@

\paragraph{Extending and combining the current challenges.}
% thanks, copilot

We have aimed to reduce the overlap of the techniques and aspects that
our challenges (linearity, scope extrusion, coinduction), despite
their common use in concurrency theory research. A subsequent round
could merge these features. We suggest starting with three extensions:
\emph{(1)} adding choice to both type and process levels, \emph{(2)}
introducing recursion and recursive types for typed systems, and
\emph{(3)} incorporating channel delegation in calculi lacking it.
%\todo{This paragraph is very long for how relatively little it says. Maybe remove/cut down?}

% While we have tried to minimise the overlap of the techniques and aspects our challenges explore (linearity, scope extrusion, coinduction), most research on concurrency theory combines some or all of these.
% A second round of formalisations might entail combining the features of the challenges.
% We consider the following three extensions to be obvious places to start:
% \emph{(1)} the addition of choice to the type level and the process level,
% \emph{(2)} the addition of recursion and (for typed systems) recursive types,
% and \emph{(3)} the addition of channel delegation to those calculi that do not already support it.%

\paragraph{Addressing new calculi and new features.}
\todo{Maybe cite some references in this section? Mention the higher-order pi calculus?}
Extending in this direction means proposing new challenges covering different features of process calculi, or different proof mechanisation features.
Some interesting aspects of message-passing calculi to be explored in further challenges could be multiparty session types and choreographies, as their meta-theory includes aspects -- \eg, well-formedness conditions on global types, partiality of end-point projection function, etc. -- that we do not address here.
Also, one could easily design a challenge that goes beyond our coinduction challenge, by exploring different notions of bisimilarity (barbed, weak, \etc) and trace equivalence.
%
\luca{Coinduction may also play a relevant role in supporting recursive
processes and session types. Indeed, it is often the case that session type
systems (and logics) with recursive session types (and propositions) are
naturally expressed in \emph{infinitary form} by interpreting their typing (and
proof) rules coinductively. I CAN ADD ONE OR TWO REFERENCES IF YOU FEEL IT'S APPROPRIATE}
